\pkg{ShinyMICE}: an Evaluation Suite for Multiple Imputation

### 2019-10-31 ResearchReport/article.Rnw
Archive answers to questions about simulation

All programming code used in this paper is available in the file XYZ.R along with the manuscript and on Github repository XYZ.

@@

@@

@@

\begin{enumerate}
\item A missingness mechanism is said to be ignorable when the probability to be missing does not depend on the missing data itself. Violation of this assumption can gravely affect inferences. Robustness of inferences to varying degrees of violation can be assessed with sensitivity analyses. Some practical guidelines exist (e.g., \citep{nguy17}), but current MI software does not facilitate this methodology for empirical researchers. 
%
\item Congeneal imputation models capture all required relations between observed and missing parts of the data. The extent to which this has been successful can be evaluated by plotting conditional distributions \citep{abay08}. Such visualizations are available in MICE, but subsequent statistical tests to quantify the relations with covariates are not provided. 
% ^[Additionally, there is potential to assess model fit by means of \emph{over-imputation} \citep{buur18} or *double robustness* \citep{bang05}-- topics that are only persued if time permits.] 
%
\item The third assumption is met when the MI algorithm converges to a stable distribution. However, conventional measures to diagnose convergence-- e.g., Gelman and Rubin's \citeyear{gelm92} statistic $\widehat{R}$ --are not applicable on multiply imputed data \citep{lace07}. Therefore, empirical researchers have to rely on visual inspection procedures that are theoretically equivalent to $\widehat{R}$ \citep{whit11}. Visually assessing convergence is not only difficult to the untrained eye, it might also be futile. The convergence properties of MI algorithms lack scientific consensus \citep{taka17}, and some default MICE techniques might not converge to stable distributions at all \citep{murr18}. Moreover, convergence diagnostics for MI methods have not been systematically studied \citep{buur18}.
\end{enumerate}

@@

@@

The\proglang{R}code and documentation of this project will be open source (available on Github). Since the study does not require the use of unpublished empirical data, I expect that the FETC will grant the label exempt.

% \includegraphics[height=80mm,keepaspectratio]{logo} % this is not correct yet, it needs a figure label and stuff

@@

@@

@@

"It is often useful to inspect the distributions of original and the imputed data. One way of doing this is to use the function \fct{stripplot} \dots  The differences between the
red points represents our uncertainty about the true (but unknown) values. \dots Under MCAR, univariate distributions of the observed and imputed data are expected to be identical. Under MAR, they can be different, both in location and spread, but their multivariate distribution is assumed to be identical." \cite[p.~12]{mice}. Create scatterplots via \fct{xyplot}.

@@

- Replicate simulation study and build a decision rule to solve the problem: "The monitoring statistic was computed for mean monthly earnings at each iteration in chains of length k = 200. Since calculation of the statistic requires M parallel sequences, m = 5 such chains were constructed. This value of m was informed by the preferred choice given in the literature on multiple imputation. The monitoring statistic computed at each iteration is presented in Figures 15 to 17 for each of the three missingness mechanisms. The red vertical line denotes ten iterations" \cite[p.~49]{lace07}. % also, use '\citep[see][for further details]{...}' for '(see Author Year for further details)'

>>>>

The aim of this paper is to provide applied researchers an introduction to the interactive evaluation device \pkg{ShinyMICE}. All programming code used in this paper is available in the file XYZ.R along with the manuscript, and on Github repository XYZ.

@@

Notation in this paper is copied from \cite{buur18}, and thus based on \cite{rubin87}. Let $Y$ denote an $n \times p$ matrix containing the data values on $p$ variables for all $n$ units in a sample. The collection of observed data values is denoted as $Y_{obs}$; the missing part of $Y$ is refered to as $Y_{mis}$. Response indicator $R$ shows whether a data value in $Y$ is missing or observed. The relation between $R$, $Y_{obs}$, and $Y_{mis}$ determines the missingness mechanism.

% "Let Y denote the n x p matrix containing the data values on p variables for all n units in the sample. We define the response indicator R as an n x p 0-1 matrix. The elements of Y and R are denoted by y i j and r i j , respectively, where i = 1 , ... , n and j = 1 , ... , p . If y i j is observed, then r i j = 1 , and if y i j is missing, then r i j = 0 ." ... The observed data are collectively denoted by Y o b s . The missing data are collectively denoted as Y m i s , and contain all elements y i j where r i j = 0 . When taken together Y = ( Y o b s , Y m i s ) contain the hypothetically complete data." \cite[par.~2.2]{buur18}

% The data are said to be MCAR if Pr ( R = 0 | Y o b s , Y m i s , $\psi$ ) = Pr ( R = 0 | $\psi$ ) (2.1) so the probability of being missing depends only on some parameters $\psi$ , the overall probability of being missing. The data are said to be MAR if Pr ( R = 0 | Y o b s , Y m i s , $\psi$ ) = Pr ( R = 0 | Y o b s , $\psi$ ) (2.2) so the missingness probability may depend on observed information, including any design factors. Finally, the data are MNAR if Pr ( R = 0 | Y o b s , Y m i s , $\psi$ ) (2.3) does not simplify, so here the probability to be missing also depends on unobserved information, including Y m i s itself. \cite[par.~2.2]{buur18}

Terminology (MCAR, MAR, MNAR).

@@

\subsection{Theoretical Background from Thesis Proposal}

@@

\begin{enumerate}
\item A missingness mechanism is said to be ignorable when the probability to be missing does not depend on the missing data itself. Violation of this assumption can gravely affect inferences. Robustness of inferences to varying degrees of violation can be assessed with sensitivity analyses. Some practical guidelines exist (e.g., \citep{nguy17}), but current MI software does not facilitate this methodology for empirical researchers.  % use '\citep[see][for further details]{...}' for '(see Author Year for further details)'
%
\item Congeneal imputation models capture all required relations between observed and missing parts of the data. The extent to which this has been successful can be evaluated by plotting conditional distributions \citep{abay08}. Such visualizations are available in MICE, but subsequent statistical tests to quantify the relations with covariates are not provided. 
% Additionally, there is potential to assess model fit by means of \emph{over-imputation} \citep{buur18} or *double robustness* \citep{bang05}-- topics that are only persued if time permits.
%
\item The third assumption is met when the MI algorithm converges to a stable distribution. However, conventional measures to diagnose convergence-- e.g., Gelman and Rubin's \citeyear{gelm92} statistic $\widehat{R}$ --are not applicable on multiply imputed data \citep{lace07}. Therefore, empirical researchers have to rely on visual inspection procedures that are theoretically equivalent to $\widehat{R}$ \citep{whit11}. Visually assessing convergence is not only difficult to the untrained eye, it might also be futile. The convergence properties of MI algorithms lack scientific consensus \citep{taka17}, and some default MICE techniques might not converge to stable distributions at all \citep{murr18}. Moreover, convergence diagnostics for MI methods have not been systematically studied \citep{buur18}.
\end{enumerate}

@@

\subsection{Approach from Thesis Proposal}

@@

The \proglang{R} code and documentation of this project will be open source (available on Github). Since the study does not require the use of unpublished empirical data, I expect that the FETC will grant the label exempt.

% \includegraphics[height=80mm,keepaspectratio]{impression} % this is not correct yet, it needs a figure label and stuff

@@

- FIMD on sensitivity analyses, see https://stefvanbuuren.name/fimd/sec-sensitivity.html

@@

- Evaluation of MI methods in SAS, see https://onlinelibrary.wiley.com/doi/full/10.1111/1467-9574.00219

@@

"The mice package contains several graphic functions that can be used to gain insight into the correspondence of the observed and imputed data: \fct{bwplot}, \fct{stripplot}, \fct{densityplot} and \fct{xyplot}." \cite[par.~6.6]{buur18}

"It is often useful to inspect the distributions of original and the imputed data. One way of doing this is to use the function \fct{stripplot} \dots  The differences between the red points represents our uncertainty about the true (but unknown) values. \dots Under MCAR, univariate distributions of the observed and imputed data are expected to be identical. Under MAR, they can be different, both in location and spread, but their multivariate distribution is assumed to be identical." \cite[p.~12]{mice}.

"Bondarenko and Raghunathan (2016) [see https://www.ncbi.nlm.nih.gov/pubmed/26952693] proposed a more refined diagnostic tool that aims to compare the distributions of observed and imputed data conditional on the missingness probability. The idea is that under MAR the conditional distributions should be similar if the assumed model for creating multiple imputations has a good fit. An example is created as ..."  \cite[par.~6.6]{buur18}

% fit <- with(imp, glm(ici(imp) ~ age + bmi + hyp + chl,
%                      family = binomial))
% ps <- rep(rowMeans(sapply(fit$analyses, fitted.values)),
%           imp$m + 1)
% xyplot(imp, bmi ~ ps | as.factor(.imp),
%        xlab = "Probability that record is incomplete",
%        ylab = "BMI", pch = c(1, 19), col = mdc(1:2))

@@

- Replicate simulation study and build a decision rule to solve the problem: "The monitoring statistic was computed for mean monthly earnings at each iteration in chains of length k = 200. Since calculation of the statistic requires M parallel sequences, m = 5 such chains were constructed. This value of m was informed by the preferred choice given in the literature on multiple imputation. The monitoring statistic computed at each iteration is presented in Figures 15 to 17 for each of the three missingness mechanisms. The red vertical line denotes ten iterations" \cite[p.~49]{lace07}.
